---
title: "Patch analysis with diamondback"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Patch analysis with diamondback}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# The vignette needs the Python backend. diamondback never provisions one on its
# own, so on a machine without NumPy and SciPy this renders as prose rather than
# quietly downloading an interpreter to build a vignette.
has_python <- tryCatch({
  diamondback::diamondback_python(quiet = TRUE); TRUE
}, error = function(e) FALSE)
knitr::opts_chunk$set(eval = has_python)
```

```{r, echo = FALSE, results = "asis", eval = !has_python}
cat("> **Note:** the code below was not run, because no Python with NumPy and",
    "SciPy was found. See `diamondback_check()` and `diamondback_install_python()`.\n")
```

```{r setup, message = FALSE}
library(diamondback)
library(terra)
```

This vignette walks through a complete workflow: label patches in a landscape,
measure them, calculate core area, and track what happens to them over time.

Everything below runs on small synthetic rasters so the answers are checkable by
eye. The whole point of the package is the same code path at 400 million cells.

## Check the environment first

diamondback runs its array kernels in NumPy and SciPy. Before a long job, make
sure the whole path works — not just that the pieces are installed:

```{r check}
diamondback_check()
```

That runs a real labelling operation end to end. If it reports success,
everything in this vignette will work.

diamondback will not install Python for you. If you already have NumPy and SciPy
somewhere it finds them; if not, `diamondback_check()` says so and you choose:
`pip install numpy scipy` into your own environment, point
`options(diamondback.python = )` at one that has them, or run
`diamondback_install_python()` to set up a private one (the only thing here that
downloads, and it asks first).

## A landscape

A small forest raster with a study-area mask, some genuinely missing data, and
30 m cells in an equal-area projection:

```{r landscape}
set.seed(42)

forest <- rast(nrows = 60, ncols = 60, xmin = 0, xmax = 1800,
               ymin = 0, ymax = 1800, crs = "EPSG:5070")
values(forest) <- rbinom(ncell(forest), 1, 0.42)
forest <- focal(forest, w = 5, fun = "mean", na.rm = TRUE)
forest <- ifel(forest > 0.45, 1, 0)

# A cloud of missing data: unknown, not absent.
forest[20:24, 20:24] <- NA

# The study area excludes a corner.
study <- rast(forest)
values(study) <- 1
study[50:60, 50:60] <- 0

plot(forest, main = "forest (1 = forest, 0 = not, NA = unknown)")
```

## Labelling

`analyze_patches()` is the function to reach for: it labels and measures in one
call, and caches the result if you give it an `output_dir`.

```{r analyze}
p1 <- analyze_patches(
  forest,
  class      = 1,
  mask       = study,
  directions = 8,
  crop       = FALSE,
  quiet      = TRUE
)

p1
```

Note what the print method leads with: the four cell states. This is the number
to check first on real data, because the most common way this analysis goes
wrong is a misclassified `NA`.

### The four states

diamondback never lets `NA` drift into foreground or background:

```{r states}
str(p1$metadata$cells)
```

Those four numbers always sum to the cell count. The labelled raster can only
show three of them — a raster has one no-data value — but `patch_domain()`
reconstructs the distinction:

```{r domain}
dom <- patch_domain(p1)
plot(dom, main = "cell states")
freq(dom)
```

`missing` is the cloud we punched in; `outside` is the masked corner. Both are
`NA` in `p1$patches`, and neither ever became a patch.

### Connectivity

`directions = 8` joins diagonal neighbours, `4` does not. This is a real
decision, not a detail:

```{r connectivity}
c(
  eight = label_patches(forest, class = 1, mask = study, directions = 8,
                        crop = FALSE, quiet = TRUE)$metadata$n_patches,
  four  = label_patches(forest, class = 1, mask = study, directions = 4,
                        crop = FALSE, quiet = TRUE)$metadata$n_patches
)
```

Under the rook rule the same landscape has more, smaller patches. Whichever you
pick, it is recorded in the metadata and it invalidates a cache built with the
other.

## Metrics

```{r metrics}
head(p1$metrics[, c("patch_id", "cells", "area_ha", "perimeter",
                    "edge_valid", "edge_domain", "touches_domain_edge")])
```

### Reading the edge columns

`perimeter` is the total length of shared cell edges between a patch and
anything that is not that patch. Corner contact contributes nothing: two cells
touching only at a corner share no physical boundary.

It always splits into two parts:

- **`edge_valid`** — boundary against a real, in-domain cell. The patch stops
  here. This is habitat edge, the thing with ecological meaning.
- **`edge_domain`** — boundary against a masked cell, a missing cell, or the
  grid border. This is an artefact of where you drew the study area.

```{r edge-sum}
with(p1$metrics, all.equal(edge_valid + edge_domain, perimeter))
```

Patches flagged `touches_domain_edge` are cut off by the study area, so their
area and perimeter are **lower bounds**:

```{r edge-flag}
table(p1$metrics$touches_domain_edge)

# The usual move: analyse the size distribution on interior patches only.
interior <- subset(p1$metrics, !touches_domain_edge)
summary(interior$area_ha)
```

Ignoring that flag is how a clipped 500-hectare patch gets counted as a small
one.

## Core area

Habitat more than a given distance from an edge, via one exact Euclidean
distance transform:

```{r core}
p1 <- patch_core_area(p1, depth = 60, units = "m", quiet = TRUE)
head(p1$metrics[, c("patch_id", "cells", "core_cells", "core_area_ha",
                    "core_fraction")])
```

Distance is measured **centre-to-centre**: from each habitat cell's centre to
the centre of the nearest non-habitat cell. This is worth being precise about,
because a nominal depth means two different things half a cell apart.

With 30 m cells, a cell sitting immediately next to non-habitat is at distance
**30 m**, not 0 and not 15 — even though the physical patch boundary is only
15 m from its centre. So `depth = 60` keeps cells with at least two cells of
habitat between them and the nearest non-habitat cell.

If you want the depth measured from the physical boundary instead, add half a
cell:

```{r core-halfcell}
centre_rule   <- patch_core_area(p1, depth = 60, quiet = TRUE)
boundary_rule <- patch_core_area(p1, depth = 60 + 15, quiet = TRUE)
c(centre = sum(centre_rule$metrics$core_cells),
  boundary = sum(boundary_rule$metrics$core_cells))
```

`edge = "background"` changes what counts as an edge — masked, missing and
off-grid cells become habitat, so patches clipped by the study boundary are not
penalised for it:

```{r core-edge}
strict <- patch_core_area(p1, depth = 60, edge = "all", quiet = TRUE)
lenient <- patch_core_area(p1, depth = 60, edge = "background", quiet = TRUE)

c(all = sum(strict$metrics$core_cells),
  background = sum(lenient$metrics$core_cells))
```

Which is right depends on your study boundary. A coastline is a real edge; an
arbitrary analytical cut through continuous forest is not.

## Change over time

A second time step: clearing cuts through the landscape.

```{r time2}
forest2 <- forest
forest2[30:33, ] <- 0    # a road-like clearing
forest2[10:14, 10:14] <- 0

p2 <- analyze_patches(forest2, class = 1, mask = study, directions = 8,
                      crop = FALSE, quiet = TRUE)

change <- compare_patches(p1, p2, quiet = TRUE)
change
```

`compare_patches()` validates that the two rasters share a CRS, extent,
resolution, origin and dimensions. It will not resample silently — a half-cell
offset would still produce a complete and entirely wrong overlap table.

### Three tables, three questions

**`$events`** answers "what happened to this patch?":

```{r events}
head(change$events[, c("time", "patch_id", "cells", "event", "n_links",
                       "primary_id", "prop_primary")])

table(change$events$event)
```

**`$overlaps`** is the raw evidence, with no lineage rule baked in:

```{r overlaps}
head(change$overlaps[, c("id1", "id2", "overlap_cells",
                         "prop_t1_retained", "prop_t2_inherited",
                         "rank_from_t1", "rank_from_t2")])
```

**`$lineages`** is one row per connected group of related patches, which is the
honest unit when patches merge:

```{r lineages}
head(change$lineages)
```

### The mother patch, and not being stuck with it

For each time-2 patch, `primary_id` is its mother: the time-1 patch that
contributed the most overlapping cells. Ties break deterministically (larger
patch, then lower ID) and are flagged:

```{r ties}
sum(change$events$primary_tie, na.rm = TRUE)
```

That default is a choice, and you are not stuck with it. `$overlaps` carries
both directional proportions, so a different rule is a re-derivation rather than
a re-run. For example, using *proportion of the time-1 patch retained* instead of
absolute overlap:

```{r alt-rule}
ov <- change$overlaps
alt <- do.call(rbind, lapply(split(ov, ov$id2), function(d) {
  d[which.max(d$prop_t1_retained), c("id2", "id1", "prop_t1_retained")]
}))
head(alt)
```

### Thresholds

By default every overlap counts, including a single incidental cell. That
default keeps the evidence complete, but classification is sensitive to it: one
shared cell is the difference between "the patch split" and "the patch
persisted".

```{r thresholds}
loose  <- compare_patches(p1, p2, min_overlap_cells = 1, quiet = TRUE)
strict <- compare_patches(p1, p2, min_overlap_prop = 0.05, quiet = TRUE)

rbind(loose = table(factor(loose$events$event,
                           levels = c("persistence", "split", "merger",
                                      "complex", "appearance", "disappearance"))),
      strict = table(factor(strict$events$event,
                            levels = c("persistence", "split", "merger",
                                       "complex", "appearance", "disappearance"))))
```

If your rasters are noisy, raise `min_overlap_prop`. diamondback reports how many
links a threshold dropped rather than dropping them quietly.

## A time series

```{r series}
forest3 <- forest2
forest3[, 30:33] <- 0    # a second clearing, crosswise

series <- track_patch_series(
  list(forest, forest2, forest3),
  time       = c(2000, 2010, 2020),
  class      = 1,
  mask       = study,
  crop       = FALSE,
  quiet      = TRUE
)

series$summary
```

Each transition is a pairwise comparison; `$transitions` stacks them all with
`time_from` and `time_to` columns. This is deliberately not a global identity
model — it is every consecutive transition, honestly labelled, which is enough
to build most lineage analyses on.

## Caching and reproducibility

Give `analyze_patches()` an `output_dir` and results are written there and reused
when the request matches:

```{r caching}
f <- file.path(tempdir(), "forest.tif")
writeRaster(forest, f, overwrite = TRUE)
d <- file.path(tempdir(), "derived_1")

first <- analyze_patches(f, class = 1, output_dir = d, quiet = TRUE)
second <- analyze_patches(f, class = 1, output_dir = d)   # a hit
```

A hit requires the *same computation*: source fingerprint, geometry, class,
mask, connectivity, NA policy and algorithm version. Change one and it recomputes
and names what differed:

```{r cache-miss}
third <- analyze_patches(f, class = 1, directions = 4, output_dir = d)
```

Metric options deliberately do **not** invalidate a labelling cache, because they
cannot change the labels.

How strongly the source is identified is your choice. By default files under
200 MB are hashed and larger ones are matched on size and modification time —
convenient, but a heuristic: a file rewritten at the same size with its timestamp
restored would look unchanged. On a pipeline whose inputs get regenerated, ask
for the stronger check:

```{r fingerprint}
strict <- analyze_patches(f, class = 1, output_dir = d, fingerprint = "full")
```

### Patch IDs are not identities

This is the easiest way to get a wrong answer with this package:

```{r ids}
head(p1$metrics$patch_id)
```

IDs are assigned in raster scan order. They are perfectly deterministic — the
same inputs always give the same numbering — but they carry no meaning across
runs. Patch 47 in 2000 and patch 47 in 2020 are unrelated, and cropping,
changing the mask, or one cell flipping near the top-left renumbers everything
after it.

Never join two runs on `patch_id`. Correspondence between times comes only from
`compare_patches()`, which derives it from actual cell overlap.

The result on disk is three plain files — a GeoTIFF, a CSV and a JSON — readable
without R or Python:

```{r files}
list.files(d)
```

Everything needed to reproduce the run is in the metadata:

```{r repro}
m <- read_patch_result(d)$metadata
str(m[c("package_version", "algorithm_version", "directions", "na", "class")])
str(m$backend)
```

## Validating

```{r validate}
validate_patch_result(p1)
```

These are independent re-derivations, not assertions restating the code: patch
counts are recounted from the raster, foreground cells recounted from the cell
states, sums compared, mask leakage checked, and edge components reconciled
against perimeter. Worth running on the first real pass of any new pipeline.

## Large rasters

Before committing to a long job:

```{r memory}
db_memory_report(forest)
```

Labelling needs the array in RAM — that is `scipy.ndimage.label()`'s
requirement. diamondback estimates the cost first and **errors before**
allocating rather than dying partway:

```{r memory-fail, error = TRUE}
label_patches(forest, class = 1, memory_limit = 1000, quiet = TRUE)
```

Practical ways to stay under the ceiling:

- pass a `mask` and leave `crop = TRUE` (the default), which shrinks the grid to
  the study area before anything is allocated;
- point `x` at a filename rather than an in-memory `SpatRaster`, so cells stream
  blockwise into the NumPy array and are never an R matrix;
- use `output_dir`, so a labelled raster is written once and never relabelled.

## Geographic rasters

Lon/lat grids are supported exactly, not approximated as planar. Cell area and
east–west extent vary with latitude, so per-row geodesic geometry is used:

```{r lonlat}
g <- rast(nrows = 20, ncols = 20, xmin = 0, xmax = 20, ymin = 30, ymax = 50,
          crs = "EPSG:4326")
set.seed(7)
values(g) <- rbinom(ncell(g), 1, 0.5)

gl <- analyze_patches(g, quiet = TRUE)

# Area agrees with terra::cellSize() to the digit.
cs <- cellSize(g, unit = "m")
c(diamondback = sum(gl$metrics$area_m2),
  cellSize    = sum(values(cs)[values(g) == 1]))
```

The one exception is `patch_core_area()`: the distance transform assumes uniform
cell spacing, which a lon/lat grid does not have. It errors rather than returning
a plausible wrong number.

```{r lonlat-core, error = TRUE}
patch_core_area(gl, depth = 1000, quiet = TRUE)
```

Project to an equal-area CRS first.

## Where to go next

- `?label_patches` — the full account of cell states and `NA` handling
- `?patch_metrics` — exactly what perimeter measures
- `?compare_patches` — event classification and lineage rules
- `DESIGN.md` in the package source — the architecture and every design decision
